---
title: "Reproducible Research: Peer Assessment 1"
author: "pszjmb1"
output: 
  html_document:
    keep_md: true
---

# Utility Functions
```{r utility}
timeprint <- function(s){
  # Prints out the given string with a datestamp
  #
  # args:
  #  s: The string to print
  s <- paste(format(Sys.time(), "%a %b %d %H:%M:%S %Y"),s,sep="> ")
  print(s)
}
```
# Load and preprocess the data
```{r getData, cache=TRUE}
if(!file.exists("data")){dir.create("data")}

## Download data
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
if(!file.exists("./data/activity.zip")){
    timeprint("Downloading data...")
    download.file(fileURL, destfile = './data/activity.zip', method="curl")
    outputMsg <- paste("Raw data downloaded from:", fileURL," ", date())    
    write(outputMsg, file = paste("./data/", "activity.zip", ".download.info.md", sep=""))
    unzip('./data/activity.zip',exdir='./data')
} else{
  con <- file("./data/activity.zip.download.info.md", "rt")  
  outputMsg <- readLines(con, 1)
}
timeprint(outputMsg)
activityData <- read.csv2("./data/activity.csv", sep=",", header=TRUE)
str(activityData)
```

# Show the mean total number of steps taken per day
## Calculate the total number of steps taken per day
```{r calcStepsPerDay, cache=TRUE}
numberOfStepsPerDay <- tapply(activityData$steps, activityData$date, sum, na.rm = TRUE)
```

## Histogram of the total number of steps taken each day
```{r doHistNumSteps, cache=TRUE}
hist(numberOfStepsPerDay)
```

## Calculate and report the mean and median of the total number of steps taken per day
```{r calcMeanMedianNumSteps, cache=TRUE}
mean(numberOfStepsPerDay)
median(numberOfStepsPerDay)
```


# What is the average daily activity pattern?



# Inputing missing values



# Are there differences in activity patterns between weekdays and weekends?
